---
title: "Tutorial 3 - Graphs"
author: "Tauana Cunha"
date: "May 2020"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    #number_sections: true
---

```{r global_rmarkdown_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#https://kbroman.org/knitr_knutshell/pages/Rmarkdown.html
```


Plots are important tools in data exploration, when you are first looking at your dataset and checking how data is distributed. It is also critical at a later stage, when you carefully design and customize figures to present your work in a talk or manuscript.

Today we will go over two sets of tools available for plotting in R: basic built-in functions, and a specific graphical package, **ggplot2** (we will talk about packages in a minute).

In [The R Graph Gallery](https://www.r-graph-gallery.com), you will find [beautiful plots](https://www.r-graph-gallery.com/all-graphs.html) that can serve as inspiration and reference for your own work. Another option are these [Top 50 ggplot2 Visualizations](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html).

# Setup

First set your working directory:
```{r eval=FALSE}
setwd("/Users/tauanajc/TEACHING/2020_STRI_introR/L3_Graphs") # Type the path to your project folder
```


# Install and load packages {#packages}

R comes with many basic functions that we have already used. Many more functionalities are available through specific packages. These are sets of new R functions developed by the R community, accompanied by their own documentation (help pages). There are thousands of R packages that span a variety of purposes, for example to better manipulate data (**dplyr**), to plot phylogenetic trees (**ggtree**), to create websites (**blogdown**). Once you have installed a package, you can load it in your R session to make the new set of functions avalaible to you.

Let's install the package that we will use later today, **ggplot2** (remember that the package name is case sensitive):
```{r eval=FALSE}
install.packages("ggplot2") # Output omitted to keep the tutorial clean
```

After installing it, the package is now in your computer and you can delete the installation command from your script. For all future projects, you only need to load your installed package at the top of the script:
```{r}
library(ggplot2)
```

If a package has never been installed, you will get an error trying to load it:
```{r eval=FALSE}
library(ggrepel)
```

```{r echo=FALSE}
message("Error in library(ggrepel) : there is no package called ‘ggrepel’")
#echo = FALSE prevents chunk from printing the R code in the html
```


# Import and manipulate a dataset

Let's use the same dataset that Lee introduced in the previous class. Remember, this is an abbreviated dataset from [Dietterich et al. 2017 Ecological Applications](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1002/eap.1573). In brief, Lee collected leaves, roots, and soils from many individuals of five plant species growing throughout a site contaminated with the heavy metals zinc, lead and cadmium. He investigated the relationships between the metal concentrations in the soil, root colonization by mycorrhizal fungi (AMF), and metal concentrations in the leaves. AMF is shown as percent of root length colonized, and metal concentrations in leaves and soils are in mg/kg.

## Read data
```{r}
heavy_metals <- read.csv(file = "Dietterich et al 2017 data for R course.csv")
```

Let's look at the column names, the first few rows of this dataset and also its structure:
```{r}
colnames(heavy_metals)
head(heavy_metals)
str(heavy_metals)
```

## Change labels and codes

To make things more clear for us, let's change the column names:
```{r}
colnames(heavy_metals) <- c("SampleID", "Species", "Mycorrhizal_Colonization",
                            "Leaves_Cadmium", "Leaves_Zinc", "Leaves_Lead",
                            "Soil_Cadmium", "Soil_Lead", "Soil_Zinc")
```

Check how they are now:
```{r}
colnames(heavy_metals)
```

Let's also change the species codes, replacing them with the genus name:
```{r}
# Create a vector with the genus names
genera <- c("Ageratina", "Deschampsia", "Eupatorium", "Minuartia", "Agrostis")
# Name the elements of this vector according to their code
names(genera) <- c("AA", "DF", "ES", "MP", "AP")
genera

# Let's try indexing the new genera object
genera[c("AA", "AA", "AA")] # We get 3 times the full genus name of code AA

# Now let's give it the entire column of species
genera[heavy_metals$Species] # We get a new vector of all the genera names according to the code!

# Use the command above to overwrite the content of the column Species
heavy_metals$Species <- genera[heavy_metals$Species]
```

Check that this worked by looking at some random lines of the dataset. Use indexing to select just a few lines, and the first 5 columns:
```{r}
heavy_metals[c(1,30,50,75,90), 1:5]
```

Before plotting, let's also change the Species column from character to factor:
```{r}
heavy_metals$Species <- as.factor(heavy_metals$Species)
str(heavy_metals)
```

Finally, let's plot some graphs!


# Basic plots with built-in functions

First we will go over some of the most common plot types with built-in functions in R.


## Histogram

This is a very standard type of plot, in which you can see the frequencies of values in your data. It automatically bins the data in ranges, but you can also customize the number of bins. To plot a histogram with base R we use the **hist** command.

Let's look at the distribution of the zinc concentrations in the leaves of all plant species:
```{r}
hist(heavy_metals$Leaves_Zinc)
```

With more bins, we can better see the distribution of zinc concentrations, take a look below.

And you probably noticed that the names of the axis look terrible! You should always change them to informative names. Same goes for the title. Any of those can also be left blank.
```{r}
hist(heavy_metals$Leaves_Zinc, breaks = 20, # Specify number of bins
     main = "", # Plot title
     xlab = "Zinc concentration in plant leaves (mg/kg)", # Axes labels
     ylab = "Frequency")
```

Much better! Remember to use the help function whenever you want to see the other arguments and options for any command:
```{r}
help(hist)
?hist
```


## Boxplot

The boxplot summarizes the data, without having all data points in the graph. The lines of the box correspond to the median, quartiles, minimum and maximum values of the dataset. The spacing between the parts of the box reflect the amount of spread and skewness in the data. If present, outliers are usually plotted as dots/circles. The command here is easy to remember: **boxplot**.

Let's get a boxplot for the exact same zinc concentrations that we used in the histogram:
```{r}
boxplot(heavy_metals$Leaves_Zinc,
        ylab = "Zinc concentration in leaves (mg/kg)")
```

A bit more information on how to interpret boxplots:

* median: middle value of the ordered set, such that 50% of the values are smaller than the median and 50% are greater
* first quartile (Q1): value below which are 25% of the elements of the ordered set 
* third quartile (Q2): value below which are 75% of the elements of the ordered set = above which are 25% of the elements
* maximum: largest number of the set (excluding outliers)
* minimum: smallest number of the set (excluding outliers)
* outliers: if present, these are values that are below Q1 or above Q3 by at least 1.5 times the interval between these two quartiles (interquantile range)

It would be more interesting to look at the distribution of zinc concentrations for each one of the individual plant species. The boxplot allows you to plot a separate box for each level of a factor:
```{r}
boxplot(formula = Leaves_Zinc ~ Species, data = heavy_metals,
        ylab = "Zinc concentration in leaves (mg/kg)",
        xlab = "")
```


## Scatter plot

Scatter plots are used to visualize the relationship between two variables. The command is simply **plot**.

Let's try a scatter plot with our soil and leaf concentrations of lead:
```{r}
plot(formula = Leaves_Lead ~ Soil_Lead, data = heavy_metals,
     ylab = "Lead concentration in leaves (mg/kg)",
     xlab = "Lead concentration in soil (mg/kg)")
```

There seems to be a positive relationship betwen soil and leaf concentrations of lead. What if we look at individual plant species? Let's use a **for loop** to get a scatter plot for each of the plants:
```{r}
for(i in 1:5){ # 5 species of plant
  each_plant = subset(heavy_metals, Species == levels(heavy_metals$Species)[i]) # subset for each species
  plot(formula = Leaves_Lead ~ Soil_Lead, data = each_plant, # plot variables
       main = levels(heavy_metals$Species)[i], # title indexed to get species name
       ylab = "Lead concentration in leaves (mg/kg)",
       xlab = "Lead concentration in soil (mg/kg)")}
```

With that we see that each plant species has a somewhat different relationship between soil and leaf concentrations of metal. **Remember that this is a small sample dataset, and not the complete dataset from Lee's paper, so our plots are merely illustrative and do not necessarily reflect real results.**


## Customization with base R

There are many other types of plots available in base R, and many more options for customization, for example colors, margins and panels with multiple plots. Below is one panel example with multiple plots and customizations. Use the help function to explore each command that is not yet familiar to you.

For more examples and details on customization of base R graphs, check the help page of the command **par** and also [this](https://rstudio-pubs-static.s3.amazonaws.com/84527_6b8334fd3d9348579681b24d156e7e9d.html) and [this](https://flowingdata.com/2012/12/17/getting-started-with-charts-in-r/) websites as starting points. Help pages and google are your best friends to customize R plots.

```{r eval=FALSE}
# output omitted to keep tutorial clean
colors() # existing colors in base R
?points # symbols
```

```{r}
palette.pals() # several new palettes added as of R v4.0.0
```

Other objects needed to create a barplot that will be used below:
```{r}
# data frame for the colors you want for each species
plant_colors <- data.frame(Species = factor(levels(heavy_metals$Species)),
                           Color = c("darkgreen", "darkorange",
                                     "deepskyblue", "firebrick", "darkviolet"),
                           Symbol = c(7,19,24,0,8))
plant_colors
```

```{r}
# Creating a function to calculate the standard error - same as last week
se <- function(x) {
  sd(x)/sqrt(length(x))}
```

```{r}
# Calculate mean and standard error of zinc concentrations for each species
for_barplot <- with(data = heavy_metals,
                    aggregate(formula = Leaves_Zinc ~ Species, FUN = mean))

colnames(for_barplot) <- c("Species", "Mean_Zinc")

for_barplot$SE_Zinc <- with(data = heavy_metals,
                            aggregate(formula = Leaves_Zinc ~ Species, FUN = se))[,2]

for_barplot
```


### Multiple customized plots

```{r fig.height=8, fig.width=8}
#pdf(file = "baseRpanel.pdf", width = 8, height = 8) # Explained below

# par = graphical parameters 
par(mfrow = c(2,2)) # mfrow: panel of 2x2 plots, ordered by row

# Histogram
hist(heavy_metals$Soil_Zinc, breaks = 20,
     main = "",
     xlab = "Zinc concentration in soil (mg/kg)")

# Boxplot
boxplot(formula = Leaves_Zinc ~ Species, data = heavy_metals,
        col = plant_colors$Color,
        xaxt = "n", # Suppress the x axis completely
        ylab = "Zinc concentration in leaves (mg/kg)",
        xlab = "",
        main = "Boxplots are better than barplots:")

text(x = 1:5, y = -23, # Position to add labels in x axis
     labels = for_barplot$Species, # Text to add
     xpd = TRUE, # Enable text outside the plot region
     srt = 15) # Angle

mtext(text = "GOOD INFO/INK RATIO",
      side = 3, # Side of graph
      col = "firebrick",
      cex = .8) # Size of text

# Scatter plot
plot(formula = Mycorrhizal_Colonization*100 ~ Soil_Zinc,
     data = heavy_metals,
     col = palette.colors(n = 5, "Dark 2"),
     pch = plant_colors$Symbol,
     ylab = "Root colonization by fungi (%)",
     xlab = "Zinc concentration in soil (mg/kg)")

mtext(text = "Different color palette just to illustrate",
      side = 4,
      col = "firebrick",
      cex = .8)

## text() - adds text to plot
## mtext() - adds text into one of the four margins of the plot

# Barplot
barplot(formula = Mean_Zinc ~ Species,
        data = for_barplot,
        col = plant_colors$Color,
        xaxt = "n",
        ylab = "Zinc concentration in leaves (mg/kg)",
        xlab = "",
        ylim = c(0,80))

text(x = c(0.7,1.9,3.1,4.3,5.5),
     y = -5,
     labels = for_barplot$Species,
     xpd=TRUE,
     srt=15)

arrows(y0 = for_barplot$Mean_Zinc-for_barplot$SE_Zinc,
       y1 = for_barplot$Mean_Zinc+for_barplot$SE_Zinc,
       x0 = c(0.7,1.9,3.1,4.3,5.5),
       x1 = c(0.7,1.9,3.1,4.3,5.5),
       angle=90,
       code=3,
       length = 0.1)

mtext(text = "BAD INFO/INK RATIO",
      side = 3,
      col = "firebrick",
      cex = .8)

par(mfrow = c(1,1)) # Set graphic device back to 1 plot only

#dev.off() # Explained below
```


## Save basic plots to file

Basic graphs can be saved with functions like **pdf**, **png**, **tiff** or **jpeg**, which open an external graphics device to save your figure. The graph will not show on your RStudio window, but it will be automatically saved in your working directory. After the **pdf** command, you list your plotting commands, followed by closing the device with **dev.off**:
```{r}
pdf(file = "baseRhist.pdf", width = 5, height = 5) # Open device

hist(heavy_metals$Soil_Zinc, breaks = 20,
     main = "Zinc concentrations (mg/kg)", xlab= "Soil")

dev.off() # Close device
```

The above was a short example with a simple graph. Now use the **pdf** and **dev.off** commands on the 4-plot panel above.


# ggplot2

Although the R built-in functions can be sufficient for many purposes, there are several graphical packages that allow more flexibility and simplicity when plotting elaborate graphs. **ggplot2** is arguably the most used R tool for creating graphs nowadays.

One major difference in the usage of **ggplot2** is the syntax. The system for adding plot elements is based on building blocks, layers that you put together to create and edit any graph. Compared to base graphics, **ggplot2** can be more verbose for simple graphics, but much less verbose for complex graphics. The data should always be in a data frame. The basic structure is: `ggplot(dataset) + geom_() + theme()`

where the first part calls an empty plot with the specified dataset, then each layer for customization is added with a **+** sign. There are many **geom_** functions available, as we will see below.

You can find information on all **ggplot2** commands in this [Reference](https://ggplot2.tidyverse.org/reference/index.html) page, and a [cheatsheet](https://ggplot2.tidyverse.org) for the most used functions.

With the package loaded, you can also call help for the new functions that came with **ggplot2**:
```{r}
?ggplot
```

Let's try all the same types of graph that we built with base R.

## Histogram

Step by step, let's see what the parts do:
```{r}
ggplot(heavy_metals) # Just an empty plot, but holds the data to be used
```

Now we specify the set of aesthetics that will be used throughout the following layers (the information is said to be *inherited* by the next commands). Aesthetic (aes) are mappings of a visual element to a specific variable, for example: the position of the data (on the x and y axes), a color pattern that follows the factor levels of another variable.

```{r}
ggplot(heavy_metals, aes(x = Leaves_Zinc)) + # Dataset and variable specified to the x axis
  geom_histogram(bins = 20, na.rm = TRUE) # Specify histogram type of plot
```


### Default themes

The **theme_** command has several default, nice-looking themes to change background colors and patterns. Also use the function **labs** to change the names of the axes:
```{r}
gghist <- ggplot(heavy_metals, aes(x = Leaves_Zinc)) +
  geom_histogram(bins = 20, na.rm = TRUE) +
  theme_minimal() + # Various default themes
  labs(x = "Zinc concentration in plant leaves (mg/kg)", y = "Frequency") # Titles

gghist
```


## Boxplot

Here we add the y aesthetics, so that we can specify both the zinc concentration and the species as variables. You can also add aesthetics not on the ggplot command, but on the **geom_** function of the specific type of plot, in which case the aesthetics are not inherited by other layers. For this particular boxplot, not inheriting the data does not matter, because none of the following layers need the data itself.
```{r}
ggplot(heavy_metals) +
  geom_boxplot(aes(y = Leaves_Zinc, x = Species), na.rm = TRUE) + # boxplot with variables in aesthetics
  theme_minimal() +
  labs(x = "", y = "Zinc concentration in plant leaves (mg/kg)")
```


### Mapping color aesthetics

Add color to the border and fillings of the boxes according to a variable (in this case, the Species):
```{r}
ggplot(heavy_metals) +
  geom_boxplot(aes(y = Leaves_Zinc, x = Species,
                   col = Species, fill = Species), na.rm = TRUE) + # new color aesthetics
  theme_minimal() +
  labs(x = "", y = "Zinc concentration in plant leaves (mg/kg)")
```

But the fill hides the value of the median. We can make the fill color transparent in the **geom_** function (fixed visual attributes are set outside of **aes**). We can also change color schemes with **scale_** functions:
```{r}
ggplot(heavy_metals) +
  geom_boxplot(aes(y = Leaves_Zinc, x = Species,
                   col = Species, fill = Species),
               na.rm = TRUE,
               alpha = 0.3) + # Transparency
  theme_minimal() +
  labs(x = "", y = "Zinc concentration in plant leaves (mg/kg)") +
  scale_fill_brewer(palette = "Dark2") + # Color palette for the fill
  scale_color_brewer(palette = "Dark2") # and for the borders
```


### Color Brewer

The website [Color Brewer](http://colorbrewer2.org) is great for finding palettes for different types of data (e.g. discrete, continuous), including options that are color-blind safe. You can pick colors manually or get the names of the palettes to use in the **scale_** functions.

Also check the help to find palettes names without going to the website:
```{r}
?scale_color_brewer
```


### Theme elements

The **theme** function allows you to edit all kinds of elements of the plot that are not the data itself, such as

* legends
* fonts of axis
* titles
* background

In this graph, we don't need the legend, so we can remove it. It would also be nice to italicize the names of the genera, and make them fit better to the graph by setting the words at an angle.
```{r}
ggbox <- ggplot(heavy_metals) +
  geom_boxplot(aes(y = Leaves_Zinc, x = Species,
                   col = Species, fill = Species),
               na.rm = TRUE,
               alpha = 0.6) +
  theme_minimal() +
  labs(x = "", y = "Zinc concentration in plant leaves (mg/kg)") +
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none", # Remove legend
        axis.text.x = element_text(face = "italic", # Make names italic
                                   size = 12, # Increase font size for x labels
                                   angle = 15)) # Angle of x labels

ggbox
```


## Scatter plot

Below we recreate the scatter plot of how lead concentration of leaves relates to concentration in the soil. With base R, we used a **for loop** to create one plot per species. Let's use the aesthetics of **ggplot2** to differentiate the points of each Species by color, in the same plot. We are also going to label specific points of interest with **geom_text**, and explore more arguments of the **theme** function:
```{r}
ggpoint <- ggplot(heavy_metals, aes(x = Soil_Lead, y = Leaves_Lead)) +
  geom_point(aes(col = Species, fill = Species), # aestethics for geom_points
             na.rm = TRUE, pch = 22, size = 3) + # attributes for points
  theme_minimal() +
  theme(axis.title = element_text(size = 11),
        legend.text = element_text(face = "italic", size = 11), # Italics in the legend
        #legend.position = "top", # Position of legend
        legend.position = "none", # Position of legend
        legend.title = element_blank()) + # Remove title of legend
  scale_fill_brewer(palette = "Dark2") +
  scale_color_brewer(palette = "Dark2") +
  geom_text(aes(label = SampleID), data = subset(heavy_metals, Leaves_Lead > 10000)) + # Label subset of points
  labs(x = "Lead concentration in soil (mg/kg)",
       y = "Lead concentration in leaves (mg/kg)")

ggpoint
```


### ggrepel and geom_jitter

In the graph above, the labels are on top of our data points, which does not look very nice. For how to move the labels away from the points, check out the package **ggrepel** and its function **geom_text_repel**. You do not need to do this now, just keep this information as a reference for the future.

In other plots where points themselves overlap, you might also want to force a tiny distance between them to improve visualization. **ggplot2** can do that with **geom_jitter**, which adds a small random variation to the actual position of your data. Look it up whenever you need it.


## Barplot

We already saw how barplots are little informative compared to other types of plots, but for the sake of the comparison with base R, let's create a barplot with **ggplot2** as well.

```{r}
ggbar <- ggplot(heavy_metals, aes(y = Leaves_Zinc, x = Species,
                         col = Species, fill = Species)) +
  geom_bar(stat = "summary", fun = "mean", na.rm = TRUE) + # Barplot with mean as summary statistics
  stat_summary(fun.data = mean_se, geom = "errorbar", na.rm = TRUE, # Add SE bars
               width = .5, col = "black") +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(size = 12, face = "italic", angle = 15),
        legend.text = element_text(face = "italic", size = 12),
        legend.position = "none") +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "", y = "Zinc concentration in leaves (mg/kg)")

ggbar
```


## Assign ggplots to objects

Plots made with **ggplot2** can be assigned to new objects, which you can use as a shortcut to keep editting the plot:
```{r}
zinc_box <- ggplot(heavy_metals) +
  geom_boxplot(aes(y = Leaves_Zinc, x = Species), na.rm = TRUE)
```

```{r}
zinc_box
```

```{r}
zinc_box +
  theme_minimal() +
  labs(x = "", y = "Zinc concentration in plant leaves (mg/kg)")
```


## Facet plots

**ggplot2** uses facets to easily display several plots of the same kind, each representing a specific level of a categorical variable. Let's take the zinc data we used in the scatter plot and produce one plot per Species, all in the same figure:
```{r, fig.height=6, fig.width=5}
ggplot(heavy_metals) +
  geom_point(aes(x = Soil_Lead, y = Leaves_Lead, fill = Species),
             na.rm = TRUE, pch = 22, size = 3) +
  theme_minimal() +
  facet_wrap(~ Species, nrow = 3) + # Wrap plots by Species
  theme(axis.title = element_text(size = 12),
        legend.position = "none",
        strip.text = element_text(size = 12, face = "italic")) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Lead concentration in leaves (mg/kg)",
       y = "Lead concentration in soil (mg/kg)")
```


## Grid of independent plots

The easiest way to combine multiple plots on a page is the **plot_grid** function from the package **cowplot**. For more detail and many more elaborate options, check this [website](https://wilkelab.org/cowplot/articles/plot_grid.html).

First, go back to the four plots we produced with **ggplot2** and assign them to object names - use gghist, ggbox, ggpoint, ggbar for the names.

**As an exercise, practice installing and loading a new package:**
(Hint: go back to the [section on packages](#packages) of the tutorial)
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(cowplot)
```

```{r}
# GIVE IT A TRY:
```

Now that you have new functions from **cowplot** available, let's combine all plots into one panel:
```{r fig.width=8, fig.height=8}
metal_plots <- plot_grid(gghist, ggbox, ggpoint, ggbar,
          nrow = 2, # Plots distributed in two rows
          align = "h", # Align plots horizontally
          axis = "l", # Side on which to align
          labels = "AUTO", # Place autmatic labels on each plot
          scale = 0.88) # Scale down the size of plots to increase white space in between

metal_plots
```


## Save ggplots to file

To save a plot to a file in your computer, use the function **ggsave** and the name of your plot object:
```{r eval=FALSE}
ggsave(filename = "ggplot2panel.pdf", plot = metal_plots,
       device = "pdf", width = 8, height = 8, units = "in")
```


# Exercises!

Practice your skills with plotting on another dataset. If you have your own data, get going with your work! Otherwise, use this cool dataset about [Himalayan Climbing Expeditions](https://github.com/tacookson/data/tree/master/himalayan-expeditions). The **members.csv** file is available to you in the zipped class material in the [Schedule](http://tauanajc.github.io/courses/2020_stri_introR/page/schedule) page of the website.

Start with reading the dataset into R, and looking at which variables are present.

```{r}
# IMPORT AND EXPLORE YOUR DATA
```

1. Plot two distinct types of graph (it's ok to try out other types that we did not see together!). Feel free to use built-in functions or ggplot2. The dataset has many interesting variables, many possible questions to explore. There is no correct answer, you can pick and choose basically anything you like. The only rule is that you have to change, in one or both of the graphs, at least the following:

* axes titles
* any font element in any part of the plot
* color scheme
* some element of the legend

```{r}
# PLOT 1
```

```{r}
# PLOT 2
```

2. Now combine these two (or more) plots into one figure:

```{r}
# PANEL WITH MULTIPLE PLOTS
```

3. Save your plot(s)

```{r}
# SAVE
```

